---
title: "Interpretable eQTL Effect Sizes Using a Log of Linear Model"
author: "Andrey A. Shabalin, John Palowitch"
date: "January 28, 2017"
output:   
  BiocStyle::html_document:
    toc: true # table of content true
    self_contained: true
vignette: >
  %\VignetteIndexEntry{ACMEeqtl Overview}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

We present a package for estimation of a new model called ACME
which respects biological understanding of eQTLs.
The model involves an additive effect of
allele count and multiplicative component random noise.\
The model is named "ACME" (Additive-Contribution, Multiplicative-Error) and is defined as

$$y_i = \log(\beta_0 + \beta_1 s_i) + Z_i^T \gamma + \epsilon_i$$

where 

* $y_i$ -- log-transformed normalized gene read count from sample $i$
* $s_i$ -- minor allele count for the SNP of sample $i$
* $Z_i$ -- Zi denote the $p \times 1$ vector of covarites for the sample $i$
* $\beta_0$ -- unknown "baseline" mean of raw expression
* $\beta_0$ -- unknown linear contribution of $s_i$ on the **raw** expression
* $\gamma$ -- unknown $p \times 1$ covariate coefficien
* $\epsilon_i$ -- noise

We estimate the model using fast iterative algorithm.\
The algorithm estimates the model which is nonlinear only with respect to $\eta = \beta_1 / \beta_0$
$$y_i = \log(1 + s_i \eta) + \log(\beta_0) + Z_i^T \gamma + \epsilon_i$$

# Getting Started

# Installation

ACMEeqtl can be installed with the following command.

```{r install, eval=FALSE}
install.packages('ACMEeqtl')
```

# Using the Package

ACMEeqtl package provides functions for analysis of a single gene-SNP pair as well as fast parallel testing of all local gene-SNP pairs.

```{r loadHidden, echo=FALSE, warning=FALSE, message=FALSE}
library(pander)
panderOptions("digits", 3)
library(ACMEeqtl)
```

```{r load}
library(ACMEeqtl)
```


## Testing a Single Gene-SNP Pair

First we generate sample gene, SNP, and set of covariates.

```{r singleInit}
# Model parameters
beta0 = 10000
beta1 = 50000

# Data dimensions
nsample = 1000
ncvrt = 19

### Data generation
### Zero average covariates
cvrt = matrix(rnorm(nsample * ncvrt), nsample, ncvrt)
cvrt = t(t(cvrt) - colMeans(cvrt))

# Generate SNPs
s = rbinom(n = nsample, size = 2, prob = 0.2)

# Generate log-normalized expression
y = log(beta0 + beta1 * s) + 
	cvrt %*% rnorm(ncvrt) + 
	rnorm(nsample)
```

We provide two equivalent functions for model estimation.

* `effectSizeEstimationR` -- fully coded in R
* `effectSizeEstimationC` -- faster version with core coded in C.

```{r singleEstim}
z1 = effectSizeEstimationR(s, y, cvrt)
z2 = effectSizeEstimationC(s, y, cvrt)

pander(rbind(z1,z2))
```

Variables `z1`, `z2` show that the estimation was done in `r z1[3]` iterations, with estimated parameters

* $\hat\beta_0$ = `r as.character(round(z1[1],1))` (true parameter is `r as.character(beta0)`)
* $\hat\beta_1$ = `r as.character(round(z1[2],1))` (true parameter is `r as.character(beta1)`)
 
## Testing All Local Gene-SNP Pairs

First we generate a eQTL dataset in filematrix format 
(see [filematrix](https://cran.r-project.org/web/packages/filematrix/index.html) package).

```{r eqtlInit}
tempdirectory = tempdir();
z = create_artificial_data(
    nsample = 100,
    ngene = 500,
    nsnp = 5000,
    ncvrt = 1,
    minMAF = 0.2,
    saveDir = tempdirectory,
    returnData = FALSE,
    savefmat = TRUE,
    savetxt = FALSE,
    verbose = FALSE)
```

In this example, we use 2 CPU cores (threads) for testing of all gene-SNP pairs within 100,000 bp.

```{r eqtlEstim}
multithreadACME(
    genefm = "gene",
    snpsfm = "snps",
    glocfm = "gene_loc",
    slocfm = "snps_loc",
    cvrtfm = "cvrt",
    acmefm = "ACME",
    cisdist = 100e+03, 
    threads = 2,
    workdir = paste0(tempdirectory,"/filematrices"),
    verbose = FALSE)
```

Now the filematrix `ACME` holds estimations for all local gene-SNP pairs.

```{r eqtlLool}
fm = fm.open(paste0(tempdirectory,"/filematrices/ACME"))
TenResults = fm[,1:10];
rownames(TenResults) = rownames(fm);
close(fm);

pander(t(TenResults))
```
